---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# midoc

<!-- badges: start -->
[![R-CMD-check](https://github.com/elliecurnow/midoc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/elliecurnow/midoc/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The Multiple Imputation DOCtor (`midoc`) R package is a decision-making system which incorporates expert, up-to-date guidance to help you choose the most appropriate analysis method when there are missing data. `midoc` will guide you through your analysis, examining the structure of the dataset to advise on whether multiple imputation (MI) is needed, and if so how to perform it. 

## Installation

You can install the development version of midoc from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("elliecurnow/midoc")
```
```{r, include=FALSE}
## Example

#This is a basic example which shows you how to solve a common problem
  
```

```{r, include=FALSE}
library(midoc)
## basic example code
```

```{r, include=FALSE}
#You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this.

#You can also embed plots, for example:

```


