---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# midoc

<!-- badges: start -->
[![R-CMD-check](https://github.com/elliecurnow/midoc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/elliecurnow/midoc/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Overview

The Multiple Imputation DOCtor (`midoc`) R package is a guidance system for analysis with missing data. It incorporates expert, up-to-date methodology to help you choose the most appropriate analysis method when there are missing data. By examining the available data and the assumed causal structure, `midoc` will advise whether multiple imputation is needed, and if so, how best to perform it. 

* `descMissData` lists missing data patterns in the specified dataset

* `exploreDAG` compares the relationships in the available data with the proposed DAG

* `checkCRA` checks complete records analysis is valid under the proposed analysis model 

* `checkMI` checks multiple imputation is valid under the proposed imputation model 

* `checkModSpec` explores the parametric specification of the imputation model 

* `proposeMI` suggests multiple imputation options based on the available data and specified imputation model

* `doMImice` performs multiple imputation based on the `proposeMI` options 

You can learn more about these commands in `vignette("midoc","midoc")`.

## Installation

You can install the development version of midoc from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("elliecurnow/midoc")
```
## Usage

```{r}
library(midoc)

head(bmi)

descMissData(y="bmi7", covs="matage mated", data=bmi, plot=TRUE)

exploreDAG(mdag=" matage -> bmi7 
                  mated -> matage 
                  mated -> bmi7 
                  sep_unmeas -> mated 
                  sep_unmeas -> r
                  pregsize -> bmi7 
                  pregsize -> bwt  
                  sep_unmeas -> bwt", 
           data=bmi)

checkCRA(y="bmi7", covs="matage mated", r_cra="r",
         mdag="   matage -> bmi7 
                  mated -> matage 
                  mated -> bmi7 
                  sep_unmeas -> mated 
                  sep_unmeas -> r
                  pregsize -> bmi7 
                  pregsize -> bwt  
                  sep_unmeas -> bwt")

checkMI(dep="bmi7", preds="matage mated pregsize", r_dep="r",
        mdag="    matage -> bmi7 
                  mated -> matage 
                  mated -> bmi7 
                  sep_unmeas -> mated 
                  sep_unmeas -> r
                  pregsize -> bmi7 
                  pregsize -> bwt  
                  sep_unmeas -> bwt")

mimod_bmi7 <- checkModSpec(formula="bmi7~matage+I(matage^2)+mated+pregsize",
                           family="gaussian(identity)", data=bmi)

miprop <- proposeMI(mimodobj=mimod_bmi7, data=bmi)

doMImice(miprop, 123, substmod="lm(bmi7 ~ matage + I(matage^2) + mated)")

```

```{r, include=FALSE}
#You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this.

#You can also embed plots, for example:

```


